
---
title: "Investigating land cover change in US Wilderness areas from 1989 to 2024"
authors: Gregg, Belote, Linck
format: html
---

**Introduction**

There are 806 wilderness areas in the United States which cover 4.5% of the country’s total land area. While maintaining the highest category of land protection, US Wilderness Areas are subject to landscape-scale changes through natural processes which may substantially alter their ecology and value to people. However, the extent of change within wilderness areas remains unknown, limiting our understanding of the present-day and future ecological value of these landscapes, including their ability to serve as core wilderness areas for larger landscapes, store carbon, and provide recreational opportunities.

The use of remotely-sensed land cover data is ubiquitous in ecological research and environmental monitoring. Previous research on US wilderness areas has used satellite-based land cover data to characterize wilderness character and quality (Carver et al. 2013). Aycrigg et al. (2022) modeled future change in wilderness areas plus a 10 km buffer under different human use, land cover change, and climate change scenarios and found the greatest land cover changes were predicted for grassland, forest, and shrublands land cover classes, and that the southwestern U.S. would undergo the greatest land cover change. A study by Adhikari and Hansen (2018) used NLCD data in greater wildland ecosystems in the United States, areas which include national parks, state and tribal lands as well as surrounding areas relevant to landscape-scale ecological integrity. In Canada, land cover data was used to develop the Park Ecological Integrity Observing System, a monitoring scheme intended to detect land cover change in near-real time across 43 national parks for management purposes (Fraser and Pouliot, 2009). Nationally, Homer et al. (2020) found extensive loss of evergreen forests and an associated increase in shrub/scrub and grassland cover between 2001 – 2016, attributing forest loss to fire, disease, and pests. Given the prevalence of these same land cover classes within US wilderness areas, with these three classes representing 75% of all CONUS wilderness area land cover in 2024, we suspect wilderness areas have likely undergone similar degrees of change, representing a significant, consequential change land cover in these areas. However, we are unaware of previous research that has investigated historical land cover change within US wilderness areas specifically, limiting our understanding of how wilderness areas are changing, their current ecological status, and whether change within wildernesses areas mirrors changes outside of these areas.

In this study, we used NLCD data to describe land cover changes across US wilderness areas over 30 years. Our goal was to quantify land cover change in wilderness areas between 1989 and 2024 and identify which wilderness areas and land cover categories have experienced the most change. Additionally, we sought to determine which US states and level two ecoregions have experienced the most change in their wilderness areas.We hypothesized that wilderness areas in more arid locations in the country have experienced the greatest degree of change, due to the prevalence of large-scale changes in these regions reported nationally over the last two decades.

**Methods**

US Wilderness Area (WA) boundaries were downloaded from Wilderness Connect (<https://www.wilderness.net/).> We downloaded US national boundaries (<https://www.wilderness.net/),> and national landcover data was downloaded at 30m\^2 resolution for the years 1989 and 2024 (<https://www.usgs.gov/).> We included all WAs occurring within the contiguous United States (CONUS), and merged and cleaned WAs containing multiple polygons and management agencies (but the same name). Of these WAs, 333 were designated after 1989, but were included under the assumption that these areas likely encompassed relatively undisturbed and ecologically intact landscapes before their designation. NLCD rasters were clipped using WA boundaries, and land cover classes were counted by summing pixels using the extract function in R version xx (R Core Team 2025) with RStudio version xx (Posit Team 2026) using the Terra package (Hijmans 2025).

We quantified change in two ways: (1) We counted change between NLCD classes for each wilderness, and at the state and level 2 ecoregion scale, and (2) For each wilderness area, we tracked land cover pixel transitions to understand landscape cover succession.

the percent land cover classes of NLCD classes of all WAs in 1989 and 2024, where change was quantified by comparing pixel differences between years for each WA, and for all areas combined. To investigate transitions between land cover classes, we created a pixel turnover matrix for each WA and across all CONUS wilderness areas.

**Results**

Three CSV files containing pixel counts were generated and are available under Data folder. All figures are based on these files.

**Figures**

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
